---
title: "01_EDA — Online Retail exploratory data analysis"
author: "Anish"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5, dpi = 110
)

library(tidyverse)
library(scales)
library(corrplot)
library(GGally)

set.seed(1)
theme_set(theme_minimal(base_size = 12))

# Encoding-safe constants for plots: avoid literal £ on systems whose locale
# is not UTF-8 (the source of the previous render error on the products plot).
pound  <- "£"
emdash <- "—"
```

# 1. Project outline

UCI Online Retail (id 352) — one year (Dec 2010 – Dec 2011) of transactions
from a UK-based online gift retailer. The project is unsupervised customer
segmentation, scoped to ISLR Ch. 12 methods (PCA, k-means, hierarchical
clustering).

`dataPreprocess.py` writes four files into `../data/`:

* `online_retail_clean.csv`     — positive-quantity, positive-price line items
* `online_retail_cancelled.csv` — cancellation invoices kept SEPARATELY (so
  net-of-returns features can be computed without throwing them away)
* `customer_features.csv`       — customer-level RFM and behavioural features
* `data_quality_summary.txt`    — sanity-check counts

This notebook does the EDA only; modelling lives in `02_PCA.Rmd` onwards.

# 2. Load and validate

```{r load}
tx        <- read_csv("../data/online_retail_clean.csv",
                      show_col_types = FALSE)
cancelled <- read_csv("../data/online_retail_cancelled.csv",
                      show_col_types = FALSE)
customer  <- read_csv("../data/customer_features.csv",
                      show_col_types = FALSE)

tx$InvoiceDate        <- as_datetime(tx$InvoiceDate)
tx$Date               <- as_date(tx$Date)
cancelled$InvoiceDate <- as_datetime(cancelled$InvoiceDate)

# Standardise CustomerID (capital ID) and clean Description encoding
# BEFORE any plot or table needs them.
tx <- tx |>
  rename_with(\(n) ifelse(n == "CustomerId", "CustomerID", n)) |>
  mutate(Description = iconv(Description, from = "", to = "UTF-8", sub = ""))

# Validation — fail fast if data shape isn't what later chunks assume.
stopifnot(
  "CustomerID" %in% names(tx),
  "CustomerID" %in% names(customer),
  !anyDuplicated(customer$CustomerID),
  nrow(customer) > 0,
  nrow(tx) > 0
)

list(
  transactions  = dim(tx),
  cancellations = dim(cancelled),
  customers     = dim(customer),
  date_range    = range(tx$InvoiceDate),
  countries     = n_distinct(tx$Country)
)
```

```{r glimpse-customer}
glimpse(customer)
```

# 3. Headline numbers — gross AND net

The pipeline now reports both gross sales and net-of-cancellations revenue.
The difference is the share of sales customers later cancelled.

```{r kpis}
gross_rev    <- sum(tx$Amount)
return_value <- sum(abs(cancelled$Amount))
net_rev      <- gross_rev - return_value

tibble(
  customers          = n_distinct(tx$CustomerID),
  invoices           = n_distinct(tx$InvoiceNo),
  cancelled_invoices = n_distinct(cancelled$InvoiceNo),
  unique_products    = n_distinct(tx$StockCode),
  line_items         = nrow(tx),
  gross_revenue      = gross_rev,
  return_value_abs   = return_value,
  net_revenue        = net_rev,
  return_value_rate  = return_value / gross_rev
) |>
  pivot_longer(everything(), names_to = "metric", values_to = "value") |>
  mutate(value = case_when(
    str_detect(metric, "rate") ~ percent(value, accuracy = 0.1),
    TRUE                       ~ comma(value, accuracy = 0.01)
  ))
```

# 4. Distribution of RFM features (raw vs. log)

```{r rfm-raw, fig.height=3.5}
customer |>
  select(Recency, Frequency, GrossMonetary) |>
  pivot_longer(everything(), names_to = "feature", values_to = "value") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 50, fill = "steelblue", colour = "white") +
  facet_wrap(~ feature, scales = "free", nrow = 1) +
  labs(title = "Raw RFM distributions (heavy right tail)",
       x = NULL, y = "Customers")
```

```{r rfm-log, fig.height=3.5}
customer |>
  select(log_Recency, log_Frequency, log_NetMonetary) |>
  pivot_longer(everything(), names_to = "feature", values_to = "value") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 50, fill = "darkorange", colour = "white") +
  facet_wrap(~ feature, scales = "free", nrow = 1) +
  labs(title = "Log-transformed RFM (closer to symmetric, but tail remains)",
       x = NULL, y = "Customers")
```

The log transform reduces — but does not eliminate — skew. `log_NetMonetary`
is approximately bell-shaped, but the upper shoulder represents real
high-value (likely wholesale-leaning) customers and will continue to
influence distance-based clustering. We address this with sensitivity
analyses in `03_clustering.Rmd` rather than removing them.

# 5. Correlation structure — Pearson and Spearman

We report both because Pearson is sensitive to outliers; Spearman (rank-based)
is robust to them. Pairs that look correlated under Pearson but not Spearman
are correlated *because of* extreme values — useful to know when choosing
which features to keep.

```{r corr-pearson, fig.height=6}
num_cols <- customer |>
  select(Recency, Frequency, GrossMonetary, NetMonetary,
         TotalQuantity, UniqueProducts,
         AvgUnitPrice_winsor, AvgOrderValue, AvgOrderItems,
         TenureDays, ReturnInvoices, ReturnRate, ReturnValueRate)

corrplot::corrplot(
  cor(num_cols, use = "pairwise.complete.obs"),
  method = "color", type = "upper", order = "hclust",
  addCoef.col = "black", number.cex = 0.6,
  tl.col = "black", tl.cex = 0.7, diag = FALSE,
  title = "Pearson", mar = c(0, 0, 2, 0)
)
```

```{r corr-spearman, fig.height=6}
corrplot::corrplot(
  cor(num_cols, use = "pairwise.complete.obs", method = "spearman"),
  method = "color", type = "upper", order = "hclust",
  addCoef.col = "black", number.cex = 0.6,
  tl.col = "black", tl.cex = 0.7, diag = FALSE,
  title = "Spearman (rank-based)", mar = c(0, 0, 2, 0)
)
```

What this implies for clustering input selection:

* `GrossMonetary`, `NetMonetary`, `TotalQuantity`, and `Frequency` form a
  tight collinear block. We keep **`NetMonetary`** because it captures the
  customer's *actual* contribution to revenue net of returns.
* `AvgOrderValue` and `AvgOrderItems` carry similar information; we keep
  `AvgOrderValue` (revenue-relevant) and drop `AvgOrderItems`.
* `Recency` is largely independent of the spending features under both
  correlation measures — it is its own axis of variation, justifying the
  canonical R, F, M triple.
* `ReturnRate` and `ReturnValueRate` are weakly correlated with everything
  else. They are best used as *profiling* variables, not clustering inputs.

# 6. Pairs plot on the modelling set

```{r pairs, fig.height=7, fig.width=8, cache=TRUE}
set.seed(1)
customer |>
  slice_sample(n = 1000) |>
  select(log_Recency, log_Frequency, log_NetMonetary,
         log_UniqueProducts, log_AvgOrderValue) |>
  GGally::ggpairs(progress = FALSE,
                  lower = list(continuous = wrap("points", alpha = 0.3,
                                                 size = 0.6))) +
  labs(title = "Log-feature pairs plot (1k-customer sample)")
```

# 7. Sales over time

```{r ts-daily}
daily <- tx |>
  group_by(Date) |>
  summarise(revenue  = sum(Amount),
            invoices = n_distinct(InvoiceNo), .groups = "drop")

ggplot(daily, aes(Date, revenue)) +
  geom_line(colour = "steelblue") +
  geom_smooth(method = "loess", span = 0.2, se = FALSE,
              colour = "darkorange") +
  scale_y_continuous(labels = label_comma(prefix = pound)) +
  labs(title = paste("Daily revenue, Dec 2010", emdash, "Dec 2011"),
       x = NULL, y = "Revenue")
```

The Q4 2011 ramp is the holiday gift season — a sanity check that the
timestamps and currency conversions are correct. The single-day spikes in
late November / early December correspond to a small number of very large
invoices (see §9 anomaly check).

```{r ts-dow-hour, fig.height=3.5}
library(gridExtra)

p_dow <- tx |>
  count(DOW = factor(DOW,
                     levels = c("Monday","Tuesday","Wednesday","Thursday",
                                "Friday","Saturday","Sunday")),
        wt = Amount, name = "revenue") |>
  ggplot(aes(DOW, revenue)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = label_comma(prefix = pound)) +
  labs(title = "Revenue by day of week", x = NULL, y = NULL)

p_hour <- tx |>
  count(Hour, wt = Amount, name = "revenue") |>
  ggplot(aes(Hour, revenue)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = label_comma(prefix = pound)) +
  labs(title = "Revenue by hour of day", x = NULL, y = NULL)

gridExtra::grid.arrange(p_dow, p_hour, nrow = 1)
```

# 8. Geography

```{r country, fig.height=4.5}
tx |>
  count(Country, wt = Amount, name = "revenue", sort = TRUE) |>
  mutate(share   = revenue / sum(revenue),
         Country = fct_reorder(Country, revenue)) |>
  slice_max(revenue, n = 10) |>
  ggplot(aes(revenue, Country)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = percent(share, accuracy = 0.1)),
            hjust = -0.1, size = 3) +
  scale_x_continuous(labels = label_comma(prefix = pound),
                     expand = expansion(mul