---
title: "02_PCA   Principal Component Analysis"
author: "Anish"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5, dpi = 110
)

library(tidyverse)
library(scales)

set.seed(1)
theme_set(theme_minimal(base_size = 12))
```

# 1. What this notebook does

PCA serves two roles in this project:

1.  **Visualisation**   project customers onto PC1–PC2 to eyeball cluster
    structure later.
2.  **Preprocessing**   replace ~7 correlated input features with 2–3
    orthogonal PC scores to feed into k-means / hclust in
    `03_clustering.Rmd`.

Output: `../data/customer_pc_scores.csv` containing
`CustomerID + PC1..PCk` (k = 3 by default).

# 2. Load and validate

```{r load}
customer <- read_csv("../data/processed/customer_features.csv",
                     show_col_types = FALSE)

stopifnot(
  "CustomerID" %in% names(customer),
  !anyDuplicated(customer$CustomerID),
  nrow(customer) > 0
)

glimpse(customer)
```

# 3. Modelling matrix

The seven-feature input drops everything that the EDA flagged as redundant
or distortive:

* `log_NetMonetary` instead of `log_GrossMonetary`   captures the
  customer's actual contribution to revenue net of returns.
* `AvgUnitPrice_winsor` instead of raw `AvgUnitPrice`   without
  winsorising, a few extreme items dominate PC2/PC3 loadings.
* `Country` excluded   UK dominance would force a near-degenerate split.
* `ReturnRate` / `ReturnValueRate` held back as profiling variables
  (sparse, mostly zero).

```{r build-X}
feature_cols <- c("log_Recency", "log_Frequency", "log_NetMonetary",
                  "log_UniqueProducts", "log_AvgOrderValue",
                  "AvgUnitPrice_winsor", "TenureDays")

X <- customer |>
  select(CustomerID, all_of(feature_cols)) |>
  drop_na()

# Sanity-check: row count must equal customer count.
stopifnot(nrow(X) == nrow(customer))

# Confirm winsorisation worked   pre-winsor max can be hundreds of pounds.
summary(X$AvgUnitPrice_winsor)
quantile(X$AvgUnitPrice_winsor, c(.5, .9, .99, 1))

dim(X)
summary(X[, feature_cols])
```

# 4. Fit PCA

`prcomp(center = TRUE, scale. = TRUE)` standardises every column to zero
mean / unit variance   required because `AvgUnitPrice_winsor` and
`TenureDays` are on totally different scales from the log features.

```{r pca-fit}
pr.out <- prcomp(X[, feature_cols], center = TRUE, scale. = TRUE)
pr.out
```

# 5. Variance explained

```{r pve}
pve <- pr.out$sdev^2 / sum(pr.out$sdev^2)
pve_df <- tibble(
  PC         = factor(paste0("PC", seq_along(pve)),
                      levels = paste0("PC", seq_along(pve))),
  PVE        = pve,
  cumulative = cumsum(pve)
)
pve_df
```

```{r scree, fig.height=3.5}
p_scree <- ggplot(pve_df, aes(PC, PVE)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = percent(PVE, accuracy = 0.1)),
            vjust = -0.4, size = 3) +
  scale_y_continuous(labels = percent_format(),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Scree plot", x = NULL, y = "Proportion of variance")

p_cum <- ggplot(pve_df, aes(PC, cumulative, group = 1)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 2) +
  geom_hline(yintercept = c(0.7, 0.8, 0.9),
             linetype = "dashed", colour = "grey60") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  labs(title = "Cumulative variance", x = NULL, y = "Cumulative PVE")

gridExtra::grid.arrange(p_scree, p_cum, nrow = 1)
```

```{r kaiser}
tibble(
  PC          = paste0("PC", seq_along(pr.out$sdev)),
  eigenvalue  = pr.out$sdev^2,
  keep_kaiser = pr.out$sdev^2 > 1
)
```

Kaiser's rule (eigenvalue > 1) gives a quick orthogonal check on the
elbow / variance-threshold heuristic. We typically keep the first three
PCs across all three rules.

# 6. Loadings   what each PC actually means

```{r loadings}
loadings <- as_tibble(pr.out$rotation, rownames = "feature")
loadings |> mutate(across(where(is.numeric), \(x) round(x, 2)))
```

```{r loadings-heatmap, fig.height=4}
loadings |>
  pivot_longer(-feature, names_to = "PC", values_to = "loading") |>
  filter(PC %in% paste0("PC", 1:4)) |>
  mutate(PC = factor(PC, levels = paste0("PC", 1:4))) |>
  ggplot(aes(PC, fct_rev(feature), fill = loading)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", loading)), size = 3) +
  scale_fill_gradient2(low = "firebrick", mid = "white",
                       high = "steelblue", midpoint = 0,
                       limits = c(-1, 1)) +
  labs(title = "Loadings (PC1–PC4)", x = NULL, y = NULL)
```

**Note on signs.** PCA component signs are arbitrary   `prcomp` could
have flipped any column. What matters is the *contrast* across features,
not the absolute sign. Read each PC as "high vs. low" along the loadings
displayed below.

Reading the loadings on the cleaned/winsorised features:

* **PC1   engagement / customer value.** Loads positively on
  `log_Frequency`, `log_NetMonetary`, `log_UniqueProducts`, and
  `TenureDays`, and negatively on `log_Recency`. High PC1 = recent,
  frequent, broad-product, repeat-relationship customers. Low PC1 =
  lapsed one-time buyers.
* **PC2   order-size vs. tenure contrast.** Strong negative loading on
  `log_AvgOrderValue` and `AvgUnitPrice_winsor`, positive on `TenureDays`
  and `log_Frequency`. Low PC2 ~ wholesale-leaning behaviour (fewer,
  larger orders of higher-priced items). High PC2 ~ smaller, more frequent
  transactions over a longer relationship. **It is *not* purely a
  "tenure / lifecycle" axis**   describing it that way ignores the
  AvgOrderValue contrast, which is the more informative half.
* **PC3   residual price-point.** Once PC1 and PC2 absorb engagement and
  order-size, PC3 picks up what's left of `AvgUnitPrice_winsor`  
  essentially the price tier the customer typically buys at.

# 7. PC scores   the new feature space

```{r scores}
scores <- as_tibble(pr.out$x) |>
  mutate(CustomerID = X$CustomerID, .before = 1) |>
  left_join(customer |> select(CustomerID, GrossMonetary, NetMonetary,
                               Frequency, Recency, Country,
                               ReturnRate, ReturnValueRate, SinglePurchase),
            by = "CustomerID")

stopifnot(nrow(scores) == nrow(X), !anyDuplicated(scores$CustomerID))
head(scores)
```

## 7a. PC1 vs PC2, coloured by net spend

```{r pc12-monetary}
ggplot(scores, aes(PC1, PC2,
                   colour = log10(pmax(NetMonetary, 1) + 1))) +
  geom_point(alpha = 0.5, size = 0.8) +
  scale_colour_viridis_c(option = "magma",
                         name = expression(log[10]~NetMonetary)) +
  labs(title = "Customers in PC1–PC2 space, coloured by net spend",
       subtitle = "PC1 should align with engagement / customer value")
```

## 7b. PC1 vs PC2, coloured by recency

```{r pc12-recency}
ggplot(scores, aes(PC1, PC2, colour = Recency)) +
  geom_point(alpha = 0.5, size = 0.8) +
  scale_colour_viridis_c(option = "viridis", direction = -1,
                         name = "Recency (days)") +
  labs(title = "PC1–PC2 space, coloured by recency",
       subtitle = "Recent customers should sit on the engagement side of PC1")
```

# 8. Biplot

A ggplot biplot reads more clearly than `biplot(pr.out)` for the report.

```{r biplot, fig.height=6, fig.width=7}
arrow_scale <- with(pr.out,
                    min(max(abs(x[, 1])) / max(abs(rotation[, 1])),
                        max(abs(x[, 2])) / max(abs(rotation[, 2])))) * 0.7

arrows_df <- as_tibble(pr.out$rotation, rownames = "feature") |>
  mutate(PC1 = PC1 * arrow_scale,
         PC2 = PC2 * arrow_scale)

ggplot(scores, aes(PC1, PC2)) +
  geom_point(alpha = 0.25, size = 0.6, colour = "grey50") +
  geom_segment(data = arrows_df,
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")),
               colour = "firebrick", linewidth = 0.7) +
  geom_text(data = arrows_df,
            aes(x = PC1 * 1.08, y = PC2 * 1.08, label = feature),
            colour = "firebrick", size = 3.5) +
  labs(title = "Biplot   customers + variable loadings on PC1–PC2",
       x = sprintf("PC1 (%s)", percent(pve[1], accuracy = 0.1)),
       y = sprintf("PC2 (%s)", percent(pve[2], accuracy = 0.1)))
```

# 9. PC1 sanity check   symmetric core, real upper tail

```{r pc1-hist, fig.height=3.5}
ggplot(scores, aes(PC1)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white") +
  labs(title = "Distribution of PC1 (engagement axis)",
       subtitle = paste("Approximately symmetric in the bulk,",
                        "but a real upper-shoulder tail of high-value customers"),
       x = "PC1", y = "Customers")
```

PC1 is approximately symmetric in the bulk but retains a right shoulder
representing the wholesale-leaning tail. We do **not** claim the
distribution is "well-behaved enough" for k-means to be unconditionally
appropriate; we instead carry this through as a known sensitivity and
report the segmentation on both raw and unit-variance-scaled PC scores
in `03_clustering.Rmd`.

```{r pc1-tail-quantiles}
quantile(scores$PC1,
         probs = c(0.5, 0.75, 0.9, 0.95, 0.99, 1.0)) |>
  round(2)
```

The gap between the 95th and 99th percentile (and again 99th to max) is
where the wholesale shoulder lives   those rows will end up in the
high-value cluster in `03_clustering.Rmd` regardless of `k`, but they
also pull the centroid of that cluster, which is why the `04_profiles.Rmd`
table reports both the median and the mean.


