---
title: "03_clustering — k-means and hierarchical"
author: "Anish"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5, dpi = 110
)

library(tidyverse)
library(scales)
library(cluster)
library(factoextra)
library(fpc)
library(mclust)        # adjustedRandIndex
library(dendextend)

set.seed(1)
theme_set(theme_minimal(base_size = 12))
```

# 1. Goal

Fit ISLR-Ch.12 clusterers (k-means and hierarchical) on the PC scores from
`02_PCA.Rmd`. Pick a `k` defensible under multiple criteria, validate
stability via bootstrap, and save per-customer labels for
`04_profiles.Rmd`.

# 2. Load and validate

```{r load}
pc       <- read_csv("../data/processed/customer_pc_scores.csv", show_col_types = FALSE)
customer <- read_csv("../data/processed/customer_features.csv",  show_col_types = FALSE)

stopifnot(
  "CustomerID" %in% names(pc),
  "CustomerID" %in% names(customer),
  !anyDuplicated(pc$CustomerID),
  setequal(pc$CustomerID, customer$CustomerID)
)

dim(pc)
glimpse(pc)
```

```{r matrix}
Xpc <- pc |> select(starts_with("PC")) |> as.matrix()
rownames(Xpc) <- pc$CustomerID
n_pc <- ncol(Xpc)

sprintf("Clustering on %d PCs across %d customers", n_pc, nrow(Xpc))
```

# 3. K-means — choosing k

## 3a. Elbow (within-cluster SSE)

The elbow plot is a heuristic: WCSS strictly decreases as `k` grows,
because more clusters can fit the data more tightly. We look for the
"kink" where adding another cluster stops paying off.

```{r km-elbow, cache=TRUE}
set.seed(1)
k_grid <- 2:10
wcss <- map_dbl(k_grid,
                \(k) kmeans(Xpc, centers = k, nstart = 25,
                            iter.max = 50)$tot.withinss)
elbow_df <- tibble(k = k_grid, wcss = wcss)

ggplot(elbow_df, aes(k, wcss)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 2) +
  scale_x_continuous(breaks = k_grid) +
  labs(title = "K-means: WCSS vs k",
       subtitle = "Look for the kink — the elbow",
       x = "k", y = "Total WCSS")
```

## 3b. Silhouette width

For each customer, silhouette width compares its average distance to
others *in its own cluster* against its average distance to the
*next-nearest cluster*. Mean silhouette ~ 0 means clusters overlap; ~ 1
means they're well separated. ISLR §12.4 doesn't formalise this, but it's
the standard add-on.

```{r km-sil, cache=TRUE}
set.seed(1)
d <- dist(Xpc)
sil <- map_dbl(k_grid, \(k) {
  cl <- kmeans(Xpc, centers = k, nstart = 25, iter.max = 50)$cluster
  mean(silhouette(cl, d)[, "sil_width"])
})
sil_df <- tibble(k = k_grid, silhouette = sil)

ggplot(sil_df, aes(k, silhouette)) +
  geom_line(colour = "darkorange", linewidth = 1) +
  geom_point(colour = "darkorange", size = 2) +
  scale_x_continuous(breaks = k_grid) +
  labs(title = "K-means: average silhouette vs k",
       x = "k", y = "Mean silhouette")
```

## 3c. Gap statistic — Tibshirani (2001)

Compares the within-cluster dispersion in the actual data against the
within-cluster dispersion under a uniform-noise reference distribution.
The **SEmax rule** picks the smallest `k` such that no larger `k` is more
than 1 standard error better — biased toward parsimony, which is exactly
what we want for an interpretable customer segmentation.

```{r km-gap, cache=TRUE}
set.seed(1)
gap <- clusGap(Xpc, FUN = kmeans, nstart = 25, K.max = 10, B = 50)
print(gap, method = "Tibs2001SEmax")

fviz_gap_stat(gap) +
  labs(title = "Gap statistic — Tibs2001SEmax rule")
```

## 3d. Decision: k = 3 (with k = 2 as sensitivity)

Diagnostics summary:

* Silhouette peaks at `k = 2` and is competitive at `k = 3`.
* Gap statistic with the SEmax rule selects `k = 3`.
* Elbow is gradual; no decisive kink.

We choose **`k = 3`** because:

1. it is statistically defensible under SEmax (the most rigorous of the
   three diagnostics);
2. the silhouette difference between `k = 2` and `k = 3` is small enough
   that gap-statistic-driven parsimony wins;
3. `k = 3` produces interpretable high / mid / low value segments rather
   than the degenerate "above-average vs. below-average" two-way split
   that `k = 2` produces;
4. the persona-level profiles in `04_profiles.Rmd` are demonstrably more
   actionable at `k = 3`.

We carry **`k = 2`** as a sensitivity / simplified segmentation. The
earlier "substantive prior of 4–6 clusters" framing is *not* what the
diagnostics support.

```{r km-final}
set.seed(1)
k_star <- 3
km_fit <- kmeans(Xpc, centers = k_star, nstart = 50, iter.max = 100)

km_fit$size
round(km_fit$centers, 2)
mean(silhouette(km_fit$cluster, d)[, "sil_width"])
```

```{r km-k2-sensitivity}
set.seed(1)
km_fit_k2 <- kmeans(Xpc, centers = 2, nstart = 50, iter.max = 100)
km_fit_k2$size
mean(silhouette(km_fit_k2$cluster, d)[, "sil_width"])
```

# 4. Sensitivity — scaling the PC scores

K-means on raw PC scores weights PC1 more heavily because PC1 has the
largest variance by construction. Scaling each PC to unit variance gives
every PC equal say in the distance metric. We check whether the
segmentation is robust to this choice using the **adjusted Rand index**
(`mclust::adjustedRandIndex`): aRI = 1 means identical labels (up to
relabelling); aRI = 0 means agreement no better than chance.

```{r scaled-pc, cache=TRUE}
set.seed(1)
km_raw    <- kmeans(Xpc,        centers = k_star, nstart = 50, iter.max = 100)
km_scaled <- kmeans(scale(Xpc), centers = k_star, nstart = 50, iter.max = 100)

aRI_scale <- adjustedRandIndex(km_raw$cluster, km_scaled$cluster)
list(adjusted_rand_raw_vs_scaled = round(aRI_scale, 3))
```

If aRI > ~0.7 the segmentation is essentially unchanged by scaling. If
materially lower, the report should note that PC1 dominance is doing the
heavy lifting and the conclusions are conditional on that choice.

# 5. Hierarchical clustering

ISLR §12.4.2. Three linkages worth comparing:

* **Ward (`ward.D2`)** — minimises within-cluster variance, philosophically
  aligned with k-means.
* **Complete** — ISLR's lab default; compact, similar-diameter clusters.
* **Average** — friendlier to elongated / unequal-density clusters; here
  as a robustness check.

```{r hc-fit, cache=TRUE}
hc_ward     <- hclust(d, method = "ward.D2")
hc_complete <- hclust(d, method = "complete")
hc_average  <- hclust(d, method = "average")
```

```{r hc-dendrograms, fig.height=10}
par(mfrow = c(3, 1), mar = c(2, 4, 3, 1))
plot(hc_ward,     labels = FALSE, hang = -1, main = "Ward (D2)",
     xlab = "", sub = "")
rect.hclust(hc_ward,     k = k_star, border = "steelblue")
plot(hc_complete, labels = FALSE, hang = -1, main = "Complete",
     xlab = "", sub = "")
rect.hclust(hc_complete, k = k_star, border = "darkorange")
plot(hc_average,  labels = FALSE, hang = -1, main = "Average",
     xlab = "", sub = "")
rect.hclust(hc_average,  k = k_star, border = "firebrick")
par(mfrow = c(1, 1))
```

```{r hc-cut}
hc_cluster <- cutree(hc_ward, k = k_star)

cross_tab <- table(kmeans = km_fit$cluster, ward = hc_cluster)
cross_tab

aRI_km_hc <- adjustedRandIndex(km_fit$cluster, hc_cluster)
list(adjusted_rand_kmeans_vs_ward = round(aRI_km_hc, 3))

mean(silhouette(hc_cluster, d)[, "sil_width"])
```

**Interpretation — be honest, not optimistic.** The cross-tab between
k-means and Ward is *not* cleanly block-diagonal at `k = 3`; substantial
off-diagonal mass means the algorithms disagree about boundaries. This is
expected: k-means prefers spherical, equal-size clusters and Ward does
not. The adjusted Rand index above quantifies