---
title: "04_profiles — naming the segments"
author: "Anish"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5, dpi = 110
)

library(tidyverse)
library(scales)
library(rpart)
library(rpart.plot)
library(knitr)

set.seed(1)
theme_set(theme_minimal(base_size = 12))

pound  <- "£"
emdash <- "—"
```

# 1. Goal

Take the k = 3 cluster labels from `03_clustering.Rmd` and turn them into
named, interpretable customer personas. Output:
`../data/customer_features_labeled.csv`.

# 2. Load + validate

```{r load}
assignments <- read_csv("../data/processed/customer_clusters.csv",
                        show_col_types = FALSE)
customer    <- read_csv("../data/processed/customer_features.csv",
                        show_col_types = FALSE)
tx          <- read_csv("../data/interim/online_retail_clean.csv",
                        show_col_types = FALSE)
cancelled   <- read_csv("../data/interim/online_retail_cancelled.csv",
                        show_col_types = FALSE)

tx$InvoiceDate        <- as_datetime(tx$InvoiceDate)
cancelled$InvoiceDate <- as_datetime(cancelled$InvoiceDate)
tx <- tx |>
  mutate(Description = iconv(Description, from = "", to = "UTF-8", sub = ""))

stopifnot(
  "CustomerID" %in% names(assignments),
  "CustomerID" %in% names(customer),
  "CustomerID" %in% names(tx),
  !anyDuplicated(customer$CustomerID),
  setequal(assignments$CustomerID, customer$CustomerID)
)

cust <- customer |>
  inner_join(assignments |> select(CustomerID, cluster),
             by = "CustomerID") |>
  mutate(cluster = factor(cluster))

list(customers_clustered = nrow(cust),
     k                   = n_distinct(cust$cluster),
     cluster_sizes       = table(cust$cluster))
```

## 2a. Re-label clusters in stable A → B → C order

K-means cluster IDs depend on the random seed. Re-label by descending
median `log_NetMonetary` so cluster `A` is always the highest-value
group, `C` the lowest, regardless of the seed.

```{r relabel}
order_by_value <- cust |>
  group_by(cluster) |>
  summarise(med = median(log_NetMonetary), .groups = "drop") |>
  arrange(desc(med)) |>
  mutate(label = LETTERS[seq_len(n())]) |>
  select(cluster, label)

cust <- cust |>
  left_join(order_by_value, by = "cluster") |>
  mutate(cluster = factor(label, levels = LETTERS[seq_along(unique(label))])) |>
  select(-label)

tx <- tx |>
  inner_join(cust |> select(CustomerID, cluster), by = "CustomerID")

table(cust$cluster)
```

# 3. Headline summary — share of customers vs share of revenue

```{r headline}
total_rev_net <- sum(cust$NetMonetary)

headline <- cust |>
  group_by(cluster) |>
  summarise(
    customers        = n(),
    revenue_net      = sum(NetMonetary),
    revenue_gross    = sum(GrossMonetary),
    median_net_spend = median(NetMonetary),
    median_orders    = median(Frequency),
    median_recency_d = median(Recency),
    .groups = "drop"
  ) |>
  mutate(pct_customers = customers   / sum(customers),
         pct_revenue   = revenue_net / total_rev_net) |>
  arrange(cluster)

headline |>
  mutate(across(c(revenue_net, revenue_gross, median_net_spend),
                \(x) dollar(x, prefix = pound)),
         across(c(pct_customers, pct_revenue),
                \(x) percent(x, accuracy = 0.1))) |>
  kable(caption = "Cluster headlines — share of customers vs revenue (net)")
```

The most important number above is **share of revenue vs share of
customers**. If cluster A holds ~28% of customers but ~75% of net
revenue, that's the segment to protect; the inverse pattern in cluster C
identifies the segment to either nurture or accept as low-margin.

```{r revenue-share, fig.height=3.5}
ggplot(headline, aes(cluster, fill = cluster)) +
  geom_col(aes(y = pct_customers), alpha = 0.4, position = "identity") +
  geom_col(aes(y = pct_revenue),   alpha = 1.0, position = "identity") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Customers (faded) vs net revenue (solid) by cluster",
       y = NULL, x = NULL) +
  theme(legend.position = "none")
```

# 4. Feature profile — median AND mean

Median is robust to outliers; mean is what aggregates revenue forecasts.
We report both because the distributions are heavy-tailed.

```{r profile-table}
profile <- cust |>
  group_by(cluster) |>
  summarise(
    n                    = n(),
    Recency_med          = median(Recency),
    Recency_mean         = mean(Recency),
    Frequency_med        = median(Frequency),
    Frequency_mean       = mean(Frequency),
    NetMonetary_med      = median(NetMonetary),
    NetMonetary_mean     = mean(NetMonetary),
    AvgOrderValue_med    = median(AvgOrderValue),
    AvgOrderValue_mean   = mean(AvgOrderValue),
    UniqueProducts_med   = median(UniqueProducts),
    UniqueProducts_mean  = mean(UniqueProducts),
    AvgUnitPrice_med     = median(AvgUnitPrice_winsor),
    TenureDays_med       = median(TenureDays),
    SinglePurchase_share = mean(SinglePurchase),
    ReturnRate_med       = median(ReturnRate),
    ReturnRate_mean      = mean(ReturnRate),
    ReturnValueRate_med  = median(ReturnValueRate),
    ReturnValueRate_mean = mean(ReturnValueRate),
    .groups = "drop"
  )

profile |>
  mutate(across(c(NetMonetary_med, NetMonetary_mean,
                  AvgOrderValue_med, AvgOrderValue_mean,
                  AvgUnitPrice_med),
                \(x) dollar(x, prefix = pound, accuracy = 0.01)),
         across(c(SinglePurchase_share,
                  ReturnRate_med, ReturnRate_mean,
                  ReturnValueRate_med, ReturnValueRate_mean),
                \(x) percent(x, accuracy = 0.1))) |>
  kable(caption = "Median and mean profile per cluster (original scales)")
```

# 5. Distributions by cluster

A median can mislead when distributions are skewed. Boxplots on the log
scale show the spread within each cluster.

```{r boxplots, fig.height=4.5}
cust |>
  select(cluster, log_Recency, log_Frequency, log_NetMonetary,
         log_AvgOrderValue, log_UniqueProducts) |>
  pivot_longer(-cluster, names_to = "feature", values_to = "value") |>
  ggplot(aes(cluster, value, fill = cluster)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.4) +
  facet_wrap(~ feature, scales = "free_y") +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Feature distributions by cluster (log scale)",
       x = NULL, y = NULL) +
  theme(legend.position = "none")
```

# 6. Parallel-coordinate signature

A single picture summarising "what makes each cluster different from the
average customer." Each line is a cluster centroid in z-scored feature
space — positive = above population mean, negative = below.

```{r signature, fig.height=4}
z_profile <- cust |>
  select(cluster, log_Recency, log_Frequency, log_NetMonetary,
         log_AvgOrderValue, log_UniqueProducts,
         AvgUnitPrice_winsor, TenureDays, ReturnValueRate) |>
  mutate(across(-cluster, \(x) as.numeric(scale(x)))) |>
  group_by(cluster) |>
  summarise(across(everything(), mean), .groups = "drop") |>
  pivot_longer(-cluster, names_to = "feature", values_to = "z")

ggplot(z_profile, aes(feature, z, group = cluster, colour = cluster)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_colour_brewer(palette = "Set1") +
  labs(title = "Cluster signature — feature means in standard deviations",
       x = NULL, y = "z-score") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

# 7. Geographic composition

`Country` was held out of clustering on purpose (UK ~ 83%). Using it now
is descriptive only — it tells us whether segments correlate with
geography after the fact.

```{r country, fig.height=4}
top_countries <- cust |> dplyr::count(Country, sort = TRUE) |>
  slice_max(n, n = 6) |> pull(Country)

cust |>
  mutate(Country = ifelse(Country %in% top_countries, Country, "Other")) |>
  dplyr::count(cluster, Country) |>
  group_by(cluster) |>
  mutate(share = n / sum(n)) |>
  ggplot(aes(cluster, share, fill = Country)) +
  geom_col() +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Country composition within each cluster",
       y = "Share of cluster", x = NULL)
```

# 8. Top products per cluster — with AND without anomaly

```{r top-products}
anomaly_code <- "23843"

top_with <- tx |>
  group_by(cluster, StockCode, Description) |>
  summarise(revenue = sum(Amount), .groups = "drop") |>
  group_by(cluster) |>
  slice_max(revenue, n = 5) |>
  ungroup()

top_with |>
  mutate(Description = str_trunc(Description, 35),
         revenue     = dollar(revenue, prefix = pound)) |>
  kable(caption = "Top 5 products by revenue, per cluster (INCLUDING 23843)")
```

```{r top-products-without}
top_without <- tx |>
  filter(StockCode != anomaly_code) |>
  group_by(cluster, StockCode, Description) |>
  summarise(revenue = sum(Amount), .groups = "drop") |>
  group_by(cluster) |>
  slice_max(revenue, n = 5) |>
  ungroup()

top_without |>
  mutate(Description = str_trunc(Description, 35),
         revenue     = dollar(revenue, prefix = pound)) |>
  kable(caption = "Top 5 products by revenue, per cluster (EXCLUDING 23843)")