---
title: "04_profiles   naming the segments"
author: "Anish"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5, dpi = 110
)

library(tidyverse)
library(scales)
library(rpart)
library(rpart.plot)
library(knitr)

set.seed(1)
theme_set(theme_minimal(base_size = 12))

pound  <- "£"
emdash <- " "
```

# 1. Goal

Take the k = 3 cluster labels from and turn them into
named, interpretable customer personas. Output:
`../data/customer_features_labeled.csv`.

# 2. Load + validate

```{r load}
assignments <- read_csv("../data/processed/customer_clusters.csv",
                        show_col_types = FALSE)
customer    <- read_csv("../data/processed/customer_features.csv",
                        show_col_types = FALSE)
tx          <- read_csv("../data/interim/online_retail_clean.csv",
                        show_col_types = FALSE)
cancelled   <- read_csv("../data/interim/online_retail_cancelled.csv",
                        show_col_types = FALSE)

tx$InvoiceDate        <- as_datetime(tx$InvoiceDate)
cancelled$InvoiceDate <- as_datetime(cancelled$InvoiceDate)
tx <- tx |>
  mutate(Description = iconv(Description, from = "", to = "UTF-8", sub = ""))

stopifnot(
  "CustomerID" %in% names(assignments),
  "CustomerID" %in% names(customer),
  "CustomerID" %in% names(tx),
  !anyDuplicated(customer$CustomerID),
  setequal(assignments$CustomerID, customer$CustomerID)
)

cust <- customer |>
  inner_join(assignments |> select(CustomerID, cluster),
             by = "CustomerID") |>
  mutate(cluster = factor(cluster))

list(customers_clustered = nrow(cust),
     k                   = n_distinct(cust$cluster),
     cluster_sizes       = table(cust$cluster))
```

## 2a. Re-label clusters in stable A → B → C order

K-means cluster IDs depend on the random seed. Re-label by descending
median `log_NetMonetary` so cluster `A` is always the highest-value
group, `C` the lowest, regardless of the seed.

```{r relabel}
order_by_value <- cust |>
  group_by(cluster) |>
  summarise(med = median(log_NetMonetary), .groups = "drop") |>
  arrange(desc(med)) |>
  mutate(label = LETTERS[seq_len(n())]) |>
  select(cluster, label)

cust <- cust |>
  left_join(order_by_value, by = "cluster") |>
  mutate(cluster = factor(label, levels = LETTERS[seq_along(unique(label))])) |>
  select(-label)

tx <- tx |>
  inner_join(cust |> select(CustomerID, cluster), by = "CustomerID")

table(cust$cluster)
```

# 3. Headline summary   share of customers vs share of revenue

```{r headline}
total_rev_net <- sum(cust$NetMonetary)

headline <- cust |>
  group_by(cluster) |>
  summarise(
    customers        = n(),
    revenue_net      = sum(NetMonetary),
    revenue_gross    = sum(GrossMonetary),
    median_net_spend = median(NetMonetary),
    median_orders    = median(Frequency),
    median_recency_d = median(Recency),
    .groups = "drop"
  ) |>
  mutate(pct_customers = customers   / sum(customers),
         pct_revenue   = revenue_net / total_rev_net) |>
  arrange(cluster)

headline |>
  mutate(across(c(revenue_net, revenue_gross, median_net_spend),
                \(x) dollar(x, prefix = pound)),
         across(c(pct_customers, pct_revenue),
                \(x) percent(x, accuracy = 0.1))) |>
  kable(caption = "Cluster headlines   share of customers vs revenue (net)")
```

The most important number above is **share of revenue vs share of
customers**. If cluster A holds ~28% of customers but ~75% of net
revenue, that's the segment to protect; the inverse pattern in cluster C
identifies the segment to either nurture or accept as low-margin.

```{r revenue-share, fig.height=3.5}
ggplot(headline, aes(cluster, fill = cluster)) +
  geom_col(aes(y = pct_customers), alpha = 0.4, position = "identity") +
  geom_col(aes(y = pct_revenue),   alpha = 1.0, position = "identity") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Customers (faded) vs net revenue (solid) by cluster",
       y = NULL, x = NULL) +
  theme(legend.position = "none")
```

# 4. Feature profile   median AND mean

Median is robust to outliers; mean is what aggregates revenue forecasts.
We report both because the distributions are heavy-tailed.

```{r profile-table}
profile <- cust |>
  group_by(cluster) |>
  summarise(
    n                    = n(),
    Recency_med          = median(Recency),
    Recency_mean         = mean(Recency),
    Frequency_med        = median(Frequency),
    Frequency_mean       = mean(Frequency),
    NetMonetary_med      = median(NetMonetary),
    NetMonetary_mean     = mean(NetMonetary),
    AvgOrderValue_med    = median(AvgOrderValue),
    AvgOrderValue_mean   = mean(AvgOrderValue),
    UniqueProducts_med   = median(UniqueProducts),
    UniqueProducts_mean  = mean(UniqueProducts),
    AvgUnitPrice_med     = median(AvgUnitPrice_winsor),
    TenureDays_med       = median(TenureDays),
    SinglePurchase_share = mean(SinglePurchase),
    ReturnRate_med       = median(ReturnRate),
    ReturnRate_mean      = mean(ReturnRate),
    ReturnValueRate_med  = median(ReturnValueRate),
    ReturnValueRate_mean = mean(ReturnValueRate),
    .groups = "drop"
  )

profile |>
  mutate(across(c(NetMonetary_med, NetMonetary_mean,
                  AvgOrderValue_med, AvgOrderValue_mean,
                  AvgUnitPrice_med),
                \(x) dollar(x, prefix = pound, accuracy = 0.01)),
         across(c(SinglePurchase_share,
                  ReturnRate_med, ReturnRate_mean,
                  ReturnValueRate_med, ReturnValueRate_mean),
                \(x) percent(x, accuracy = 0.1))) |>
  kable(caption = "Median and mean profile per cluster (original scales)")
```

# 5. Distributions by cluster

A median can mislead when distributions are skewed. Boxplots on the log
scale show the spread within each cluster.

```{r boxplots, fig.height=4.5}
cust |>
  select(cluster, log_Recency, log_Frequency, log_NetMonetary,
         log_AvgOrderValue, log_UniqueProducts) |>
  pivot_longer(-cluster, names_to = "feature", values_to = "value") |>
  ggplot(aes(cluster, value, fill = cluster)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.4) +
  facet_wrap(~ feature, scales = "free_y") +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Feature distributions by cluster (log scale)",
       x = NULL, y = NULL) +
  theme(legend.position = "none")
```

# 6. Parallel-coordinate signature

A single picture summarising "what makes each cluster different from the
average customer." Each line is a cluster centroid in z-scored feature
space   positive = above population mean, negative = below.

```{r signature, fig.height=4}
z_profile <- cust |>
  select(cluster, log_Recency, log_Frequency, log_NetMonetary,
         log_AvgOrderValue, log_UniqueProducts,
         AvgUnitPrice_winsor, TenureDays, ReturnValueRate) |>
  mutate(across(-cluster, \(x) as.numeric(scale(x)))) |>
  group_by(cluster) |>
  summarise(across(everything(), mean), .groups = "drop") |>
  pivot_longer(-cluster, names_to = "feature", values_to = "z")

ggplot(z_profile, aes(feature, z, group = cluster, colour = cluster)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_colour_brewer(palette = "Set1") +
  labs(title = "Cluster signature   feature means in standard deviations",
       x = NULL, y = "z-score") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

# 7. Geographic composition

`Country` was held out of clustering on purpose (UK ~ 83%). Using it now
is descriptive only   it tells us whether segments correlate with
geography after the fact.

```{r country, fig.height=4}
top_countries <- cust |> dplyr::count(Country, sort = TRUE) |>
  slice_max(n, n = 6) |> pull(Country)

cust |>
  mutate(Country = ifelse(Country %in% top_countries, Country, "Other")) |>
  dplyr::count(cluster, Country) |>
  group_by(cluster) |>
  mutate(share = n / sum(n)) |>
  ggplot(aes(cluster, share, fill = Country)) +
  geom_col() +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Country composition within each cluster",
       y = "Share of cluster", x = NULL)
```

# 8. Top products per cluster   with AND without anomaly

```{r top-products}
anomaly_code <- "23843"

top_with <- tx |>
  group_by(cluster, StockCode, Description) |>
  summarise(revenue = sum(Amount), .groups = "drop") |>
  group_by(cluster) |>
  slice_max(revenue, n = 5) |>
  ungroup()

top_with |>
  mutate(Description = str_trunc(Description, 35),
         revenue     = dollar(revenue, prefix = pound)) |>
  kable(caption = "Top 5 products by revenue, per cluster (INCLUDING 23843)")
```

```{r top-products-without}
top_without <- tx |>
  filter(StockCode != anomaly_code) |>
  group_by(cluster, StockCode, Description) |>
  summarise(revenue = sum(Amount), .groups = "drop") |>
  group_by(cluster) |>
  slice_max(revenue, n = 5) |>
  ungroup()

top_without |>
  mutate(Description = str_trunc(Description, 35),
         revenue     = dollar(revenue, prefix = pound)) |>
  kable(caption = "Top 5 products by revenue, per cluster (EXCLUDING 23843)")
```

The two tables together tell the anomaly story directly: if `23843`
appears as a single cluster's top product in the first table but
disappears entirely from the second, that cluster's "top product" was
really a single bulk wholesale invoice (see §9 of `01_EDA.Rmd`). The
second table is the more honest description of the segment's habitual
product mix.

# 9. What separates the clusters?   a 1-page decision tree

Centroids and parallel coordinates show *where* clusters sit. A small
classification tree shows *which features and which thresholds* you
would actually use to assign a new customer to a cluster. We fit
`rpart` on the original (un-PCA'd) features so the splits are
interpretable in business units (£, days, counts) rather than PC
scores.

```{r rpart-tree, fig.height=5}
tree_data <- cust |>
  select(cluster,
         Recency, Frequency, NetMonetary,
         AvgOrderValue, UniqueProducts,
         AvgUnitPrice_winsor, TenureDays)

tree <- rpart(cluster ~ ., data = tree_data,
              method = "class",
              control = rpart.control(cp = 0.01, maxdepth = 4,
                                      minbucket = 50))

rpart.plot(tree,
           type = 2, extra = 104, fallen.leaves = TRUE,
           box.palette = "RdYlGn", branch.lty = 3,
           main = "What separates the clusters? (rpart, depth ≤ 4)")
```

```{r rpart-accuracy}
pred <- predict(tree, tree_data, type = "class")
acc  <- mean(pred == tree_data$cluster)

list(
  in_sample_accuracy = round(acc, 3),
  confusion          = table(predicted = pred, actual = tree_data$cluster)
)
```

The tree is descriptive, not predictive: the labels themselves came
from k-means, so the in-sample accuracy is an upper bound on how
recoverable the clustering is from the original features. A tree
accuracy above ~0.85 means the clusters can be substantially explained
by 3–4 simple rules on the raw features, which is the precondition for
naming them in §10.

# 10. Persona names

The persona names below are the human-readable labels that go on
dashboards, briefs, and CRM segments. They are derived directly from
the §4 profile table and the §9 tree splits   no inputs that aren't
already in this notebook.

| Cluster | Persona | One-line description |
|---|---|---|
| **A** | **Wholesale-leaning regulars** | High net spend, frequent orders, broadest product range, often higher AvgOrderValue. Small share of customers, large share of revenue. |
| **B** | **Engaged retail** | Recent and reasonably frequent buyers at typical retail order sizes. Largest share of customers, moderate share of revenue. |
| **C** | **Lapsed / one-time** | High recency (haven't bought in a long time), low frequency, often `SinglePurchase = TRUE`. Large share of customers, small share of revenue. |

These names are consistent across the assignment-stable sense of §6 of
`03_clustering.Rmd`: persona **A** corresponds to the high-Jaccard
high-value cluster and is safe to act on; persona **B** sits in the
fuzzier middle band and should be targeted with broader campaigns;
persona **C** is presented as "directional only" wherever the bootstrap
Jaccard came in below 0.75, because the boundary between **B** and **C**
is the part of the partition that moves most under resampling.

```{r attach-persona}
persona_lookup <- tibble(
  cluster = factor(c("A", "B", "C"),
                   levels = LETTERS[seq_along(levels(cust$cluster))]),
  persona = c("Wholesale-leaning regulars",
              "Engaged retail",
              "Lapsed / one-time")
) |>
  filter(cluster %in% levels(cust$cluster))

cust <- cust |>
  left_join(persona_lookup, by = "cluster")

table(cust$cluster, cust$persona)
```

# 11. Executive summary

Three numbers the executive summary should lead with:

```{r exec-summary}
exec <- cust |>
  group_by(cluster, persona) |>
  summarise(
    customers     = n(),
    pct_customers = n() / nrow(cust),
    revenue_net   = sum(NetMonetary),
    pct_revenue   = sum(NetMonetary) / sum(cust$NetMonetary),
    median_recency_d = median(Recency),
    .groups = "drop"
  ) |>
  arrange(cluster)

exec |>
  mutate(revenue_net   = dollar(revenue_net,  prefix = pound, accuracy = 1),
         pct_customers = percent(pct_customers, accuracy = 0.1),
         pct_revenue   = percent(pct_revenue,   accuracy = 0.1)) |>
  kable(caption = "Executive summary   customers, revenue, recency by persona")
```

The recommended actions follow directly from the customer-share /
revenue-share asymmetry:

* **Persona A   protect.** Disproportionate revenue from a small
  customer base. Lifecycle programmes, account-management touchpoints,
  early access to new SKUs. Loss aversion: the cost of churning one A
  is the cost of acquiring many Bs.
* **Persona B   convert.** Largest customer base, moderate revenue
  per head. Treatment effect leverage is highest here   an offer that
  shifts a B toward A behaviour is worth more in expectation than the
  same offer to an A. Test promotions, cross-sell adjacencies, and
  basket-size nudges.
* **Persona C   selective re-engagement.** Don't blanket-target a
  segment with low Jaccard stability. Run a controlled win-back to the
  most recently lapsed slice of C and measure incrementality before
  scaling.


