---
title: "UCI Online Retail   Customer Segmentation"
subtitle: "Final report"
author: "Anish Ranjan"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
geometry: margin=0.9in
fontsize: 10.5pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", dpi = 150
)

library(tidyverse)
library(scales)
library(knitr)
library(gridExtra)

source(here::here("src", "R", "00_setup.R"))

theme_set(theme_minimal(base_size = 9))
pound <- "£"

# Pipeline outputs   produced by `make model`.
customer  <- readr::read_csv(abs_path(CONFIG$data$customer_features),
                             show_col_types = FALSE)
clusters  <- readr::read_csv(abs_path(CONFIG$data$customer_clusters),
                             show_col_types = FALSE)
pc_scores <- readr::read_csv(abs_path(CONFIG$data$customer_pc_scores),
                             show_col_types = FALSE)
pca       <- readRDS(abs_path(CONFIG$models$pca_model))

cust <- customer |>
  inner_join(clusters |> select(CustomerID, cluster), by = "CustomerID") |>
  mutate(cluster = factor(cluster, levels = LETTERS[seq_len(CONFIG$clustering$k_star)]))

scores <- pc_scores |>
  inner_join(clusters |> select(CustomerID, cluster), by = "CustomerID") |>
  inner_join(customer |> select(CustomerID, NetMonetary, Recency),
             by = "CustomerID") |>
  mutate(cluster = factor(cluster, levels = LETTERS[seq_len(CONFIG$clustering$k_star)]))
```

# Abstract

We build an unsupervised customer-segmentation pipeline on one year of
UCI Online Retail transactions [7] (~4,300 customers, ~391k positive
line items, UK gift retailer, Dec 2010 – Dec 2011). The pipeline reduces
seven engineered RFM-and-behavioural features to three principal
components (~80% variance retained) and partitions customers with
k-means at $k = 3$ [1], chosen by the gap statistic [2] and validated by
silhouette width [3], adjusted Rand index against Ward hierarchical
clustering [4], and bootstrap Jaccard stability [5]. The headline
finding is the **revenue–customer asymmetry**: cluster A (~28% of
customers) holds ~75% of net revenue, defining where retention spend
should go and which customers can be treated as low-margin volume. Next
steps are time-aware features, a soft-assignment Gaussian mixture model
benchmark, and a controlled marketing experiment per segment.

# Overview

**Problem statement.** A retailer that treats every customer the same
overspends on low-value buyers and underspends on the few that drive
revenue. The business question   *who do we protect, who do we grow,
who do we accept as low-margin?*   needs a small, defensible vocabulary
tied to behaviour rather than gut feel.

**Proposed methodology.** Per-customer feature engineering ->
column-wise standardisation -> PCA for dimensionality reduction ->
k-means on PC scores with $k$ selected by gap-statistic SE-max ->
multi-criterion validation (silhouette, ARI vs Ward, bootstrap
Jaccard) -> A/B/C re-labelling by descending median net spend ->
persisted scoring artefact for new customers.

# Data Processing

Cleaning (`src/python/data_prep.py`) drops missing-customer rows,
deduplicates, removes non-product stock codes (postage, bank charges),
**separates cancellation invoices into a sibling table** rather than
discarding them (so net monetary can be computed honestly), and
retains positive-quantity / positive-price lines in the main
table. After cleaning: ~391k positive line items across ~4,334
customers plus a ~9k cancellations table.

**Issues and adjustments.** (i) `AvgUnitPrice` is computed
quantity-weighted rather than as a line-level mean, which would
overweight tiny low-priced items. (ii) `AvgUnitPrice` is winsorised at
its 99th percentile so a few extreme prices do not dominate
PC2/PC3. (iii) Right-skewed features get $\log(1+x)$; `NetMonetary`,
which can be slightly negative, gets the sign-preserving variant
$\mathrm{sgn}(x)\log(1+|x|)$. (iv) StockCode `23843` is a single bulk
order whose product-level revenue dwarfs others; product-mix tables
are reported both with and without it.

# Data Analysis

Per-customer features: Recency, Frequency, GrossMonetary,
NetMonetary, TotalQuantity, UniqueProducts, AvgOrderValue,
AvgUnitPrice (winsorised), TenureDays, plus profiling-only fields
(Country, ReturnRate, ReturnValueRate, SinglePurchase). Two patterns
in the EDA shape every modelling decision:

```{r eda-distribution, fig.width=7, fig.height=2.4}
p_dist <- customer |>
  ggplot(aes(NetMonetary)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white") +
  scale_x_continuous(trans = "log1p",
                     labels = label_number(prefix = pound, big.mark = ",")) +
  labs(title = "Net monetary value per customer (log scale)",
       x = NULL, y = "Customers")

revenue_share <- customer |>
  arrange(desc(NetMonetary)) |>
  mutate(rank_pct = row_number() / n(),
         cum_rev  = cumsum(NetMonetary) / sum(NetMonetary))

p_concentration <- ggplot(revenue_share, aes(rank_pct, cum_rev)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_vline(xintercept = c(0.05, 0.25), linetype = "dashed",
             colour = "grey60") +
  scale_x_continuous(labels = percent_format()) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Cumulative net revenue vs. ranked customer share",
       x = "Top X% of customers (by spend)", y = "Cumulative revenue")

gridExtra::grid.arrange(p_dist, p_concentration, nrow = 1)
```

**Heavy right-skew on spend.** Net monetary spans four orders of
magnitude   distance-based methods are dominated by the upper tail
without a log transform. **Revenue is highly concentrated.** The top
5% of customers account for ~half of net revenue and the top 25% for
the great majority. This is the empirical shape the segmentation must
respect: separating the high-value tail from the low-value mass is
the minimum criterion for a useful partition.

# Model Training

**Feature engineering for the model matrix.** Seven features survive
EDA-driven pruning (collinear blocks collapsed, profiling-only fields
held back): `log_Recency`, `log_Frequency`, `log_NetMonetary` (sign-
preserving variant, since net monetary can be slightly negative),
`log_UniqueProducts`, `log_AvgOrderValue`, `AvgUnitPrice_winsor`, and
`TenureDays`. Columns are centred and scaled to unit variance.

**Evaluation metrics for $k$ selection.** (i) Within-cluster sum of
squares (elbow heuristic). (ii) Mean silhouette width [3]: $s(i)$
contrasts the within-cluster tightness $a(i)$ against the
nearest-other-cluster distance $b(i)$. (iii) Gap statistic [2] with
the SE-max rule, biased toward parsimony. We sweep $k \in \{2,
\dots, 10\}$ with `nstart = 50` and a fixed seed.

```{r pca-scree, fig.width=6.6, fig.height=2.4}
pve <- pca$sdev^2 / sum(pca$sdev^2)
pve_df <- tibble(
  PC  = factor(paste0("PC", seq_along(pve)),
               levels = paste0("PC", seq_along(pve))),
  PVE = pve, cum = cumsum(pve)
)

p_scree <- ggplot(pve_df, aes(PC, PVE)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = percent(PVE, 0.1)), vjust = -0.4, size = 2.5) +
  scale_y_continuous(labels = percent_format(),
                     expand = expansion(mult = c(0, 0.18))) +
  labs(title = "PCA scree", x = NULL, y = "PVE")

p_cum <- ggplot(pve_df, aes(PC, cum, group = 1)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 2) +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "grey60") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  labs(title = "Cumulative variance (dashed = 80%)",
       x = NULL, y = NULL)

gridExtra::grid.arrange(p_scree, p_cum, nrow = 1)
```

**Model selection.** Three principal components retain ~80% of the
variance and the elbow flattens after PC3   adding a fourth buys little
variance and costs interpretability. PC1 is the engagement / customer-
value axis (positive on $F$, $M^{\text{net}}$, $U$, $T$; negative on
$R$); PC2 is an order-size vs tenure contrast (negative on $A$, $P$);
PC3 is the residual price-point axis. For $k$, silhouette peaks at
$k = 2$ and remains competitive at $k = 3$; the gap statistic SE-max
rule selects $k = 3$. We adopt **$k = 3$** as the canonical
segmentation; $k = 2$ is retained as a sensitivity model.

# Model Validation

Three independent validations confirm the partition is real, not an
artefact of the algorithm.

```{r pca-pictures, fig.width=7.2, fig.height=3.4}
arrow_scale <- with(pca,
                    min(max(abs(scores$PC1)) / max(abs(rotation[, 1])),
                        max(abs(scores$PC2)) / max(abs(rotation[, 2])))) * 0.7

arrows_df <- as_tibble(pca$rotation, rownames = "feature") |>
  mutate(PC1 = PC1 * arrow_scale,
         PC2 = PC2 * arrow_scale,
         feature = str_remove_all(feature, "log_|_winsor"))

p_color <- ggplot(scores,
                  aes(PC1, PC2, colour = log10(pmax(NetMonetary, 1) + 1))) +
  geom_point(alpha = 0.55, size = 0.6) +
  scale_colour_viridis_c(option = "magma",
                         name = expression(log[10]~Net~spend)) +
  labs(title = "PC1 vs PC2   coloured by net spend") +
  theme(legend.key.height = unit(0.7, "cm"),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6))

p_biplot <- ggplot(scores, aes(PC1, PC2)) +
  geom_point(alpha = 0.18, size = 0.4, colour = "grey50") +
  geom_segment(data = arrows_df,
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.18, "cm")),
               colour = "firebrick", linewidth = 0.55) +
  geom_text(data = arrows_df,
            aes(x = PC1 * 1.10, y = PC2 * 1.10, label = feature),
            colour = "firebrick", size = 2.6) +
  labs(title = "Biplot   feature loadings on PC1/PC2")

gridExtra::grid.arrange(p_color, p_biplot, nrow = 1)
```

**(i) PC1 captures spend.** The colour gradient on the left panel
moves cleanly along PC1   the principal axis recovered from variance
maximisation aligns with the business-meaningful spend axis. The
biplot confirms: `Frequency`, `NetMonetary`, `UniqueProducts`, and
(negatively) `Recency` all load along PC1, while `AvgOrderValue` and
`AvgUnitPrice` load on PC2 and define wholesale-vs-retail.

**(ii) Bootstrap Jaccard stability** [5]. Resampling customers with
replacement and refitting, every k-means cluster recovers above the
0.75 stability threshold (~0.87+); Ward at the same $k$ produces one
stable cluster and two borderline ones. **(iii) ARI** [4] between
k-means and Ward is in the moderate range   the cluster *cores* are
shared, the *boundaries* differ. K-means is promoted as the canonical
label source on the grounds of stability.

**Risks and biases.** PC1's larger variance weights it more heavily
under Euclidean distance, so the segmentation is partially driven by
that axis; the raw-vs-scaled-PC ARI sensitivity reports how much that
matters in practice. The high-value tail pulls cluster A's centroid;
re-fitting with the top 1% removed is a recommended robustness check.

# Model Performance

```{r persona-table}
total_net <- sum(cust$NetMonetary)
persona_lookup <- tibble(
  cluster = factor(c("A", "B", "C"), levels = c("A", "B", "C")),
  persona = c("Loyal high-value", "Developing mid-value", "Lapsed one-time")
)
persona_summary <- cust |>
  group_by(cluster) |>
  summarise(
    customers   = n(),
    pct_cust    = n() / nrow(cust),
    pct_revenue = sum(NetMonetary) / total_net,
    median_net  = median(NetMonetary),
    median_freq = median(Frequency),
    median_rec  = median(Recency),
    .groups = "drop"
  ) |>
  left_join(persona_lookup, by = "cluster") |>
  select(cluster, persona, customers, pct_cust, pct_revenue,
         median_net, median_freq, median_rec)

persona_summary |>
  mutate(
    customers   = comma(customers),
    pct_cust    = percent(pct_cust,    accuracy = 0.1),
    pct_revenue = percent(pct_revenue, accuracy = 0.1),
    median_net  = dollar(median_net,   accuracy = 1, prefix = pound),
    median_rec  = sprintf("%d d", as.integer(median_rec))
  ) |>
  rename(`Cluster` = cluster, `Persona` = persona,
         `Customers` = customers, `% cust.` = pct_cust,
         `% rev.` = pct_revenue, `Median net spend` = median_net,
         `Med. orders` = median_freq, `Med. recency` = median_rec) |>
  kable(align = "lllrrrrr",
        caption = "Per-cluster headline numbers (live from the pipeline output)")
```

**Benchmark   segmentation quality.** Bootstrap Jaccard ≥ 0.75 across
all three k-means clusters (well above the conventional [5] cut-off);
ARI vs Ward in the moderate range; the silhouette at $k = 3$ within
~0.02 of $k = 2$ while delivering an extra interpretable segment.

**Baseline comparison.** A naive *RFM-quintile* baseline (rank
customers into 5×5×5 = 125 cells, group cells by hand) recovers the
same A/B/C ranking on aggregate revenue but produces a coarse,
manually-tuned partition with no statistical defence of cell
boundaries. K-means on PCA improves on this baseline along three axes:
(i) data-driven boundaries instead of hand-set quantile cuts,
(ii) bootstrap-validated stability rather than analyst judgment, and
(iii) a deployable scoring function that assigns new customers
without re-binning.

```{r signature-plot, fig.width=7, fig.height=2.4}
z_profile <- cust |>
  select(cluster, log_Recency, log_Frequency, log_NetMonetary,
         log_AvgOrderValue, log_UniqueProducts,
         AvgUnitPrice_winsor, TenureDays) |>
  mutate(across(-cluster, \(x) as.numeric(scale(x)))) |>
  group_by(cluster) |>
  summarise(across(everything(), mean), .groups = "drop") |>
  pivot_longer(-cluster, names_to = "feature", values_to = "z") |>
  mutate(feature = str_remove_all(feature, "log_|_winsor"))

ggplot(z_profile, aes(feature, z, group = cluster, colour = cluster)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_line(linewidth = 1) +
  geom_point(size = 2.4) +
  scale_colour_brewer(palette = "Set1", name = "Cluster") +
  labs(title = "Cluster signature   feature means in standard deviations",
       x = NULL, y = "z-score") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

**Domain reading.** **A   Loyal high-value** sits above zero on every
engagement feature; the segment to *protect*. **B   Developing
mid-value** sits above zero on tenure and frequency but at typical
retail order sizes; the segment with the highest treatment-effect
leverage   *grow*. **C   Lapsed one-time** sits below zero on every
engagement feature with strong positive Recency; the segment to
*triage* via controlled win-back, not blanket-targeted.

# Conclusion

**Positive results.** Three actionable, validated customer segments
shipped as a reproducible pipeline; the customer-share / revenue-share
asymmetry recovers a finding the business can immediately operate on.
**Negative results.** PC1 dominance is real and the segmentation is
partially scale-dependent (mitigated by the raw-vs-scaled-PC ARI
report). Ward hierarchical clustering produces two borderline clusters
at the same $k$, so cross-method agreement is partial.
**Recommendations.** Protect cluster A through retention and
account-management spend; convert cluster B with measured-uplift
promotions; triage cluster C with controlled win-back rather than
blanket campaigns. **Caveats.** The segmentation is descriptive, not
predictive; it depends on the net-vs-gross monetary choice, the
winsorisation cut-off, and the one-year window. The recommended
actions are testable hypotheses requiring marketing experiments to
confirm.

# Data Sources

UCI Machine Learning Repository, "Online Retail" (id 352), Chen et
al. [7]: <https://archive.ics.uci.edu/dataset/352/online+retail>.
Programmatic access via the `ucimlrepo` Python package (`fetch_ucirepo(id=352)`),
called from `src/python/data_prep.py`. The download is cached in
`data/raw/`; cleaned line items in `data/interim/`; engineered
customer-level features and pipeline outputs in `data/processed/`.

# Source Code

Repository layout (top-level): `src/python/data_prep.py` (cleaning +
features), `src/R/00_setup.R … 06_score_new_customers.R` (modelling
modules), `scripts/run_pipeline.R` (orchestration),
`scripts/render_reports.R` (reports), `tests/` (`testthat` integrity
checks), `reports/` (this report and the four per-stage Rmds),
`models/segmentation_pipeline.rds` (deployable artefact).
**Build:** `./setup.sh` from a clean checkout   creates the venv,
installs Python and R deps, runs `make all`. Targets: `make data` /
`model` / `reports` / `test` / `clean`.
**Dependencies (open source):** Python   `ucimlrepo`, `pandas`,
`numpy`, `PyYAML`. R   `tidyverse`, `cluster`, `factoextra`, `fpc`,
`mclust`, `dendextend`, `rpart`, `rpart.plot`, `corrplot`, `GGally`,
`gridExtra`, `knitr`, `rmarkdown`, `testthat`, `here`, `yaml`.

# Bibliography

[1] J. A. Hartigan and M. A. Wong, "Algorithm AS 136: A K-Means
Clustering Algorithm," *J. Royal Statistical Society. Series C
(Applied Statistics)*, vol. 28, no. 1, pp. 100–108, 1979.

[2] R. Tibshirani, G. Walther, and T. Hastie, "Estimating the number
of clusters in a data set via the gap statistic," *J. Royal
Statistical Society. Series B*, vol. 63, no. 2, pp. 411–423, 2001.

[3] P. J. Rousseeuw, "Silhouettes: A graphical aid to the
interpretation and validation of cluster analysis," *J. Computational
and Applied Mathematics*, vol. 20, pp. 53–65, 1987.

[4] L. Hubert and P. Arabie, "Comparing partitions," *J.
Classification*, vol. 2, no. 1, pp. 193–218, 1985.

[5] C. Hennig, "Cluster-wise assessment of cluster stability,"
*Computational Statistics & Data Analysis*, vol. 52, no. 1, pp.
258–271, 2007.

[6] G. James, D. Witten, T. Hastie, and R. Tibshirani, *An
Introduction to Statistical Learning with Applications in R*, 2nd ed.
New York: Springer, 2021.

[7] D. Chen, S. L. Sain, and K. Guo, "Data mining for the online
retail industry: A case study of RFM model-based customer segmentation
using data mining," *J. Database Marketing & Customer Strategy
Management*, vol. 19, no. 3, pp. 197–208, 2012.