---
title: "01_eda"
author: "Anish"
date: "2026-04-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  exo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5, dpi = 110
)

library(tidyverse)
library(scales)
library(corrplot)
library(GGally)

theme_set(theme_minimal(base_size = 12))
```

# 1. Project Outline

We are working with the **UCI Online Retail** dataset (id 352): roughly one year of transactions from a UK-based online gift retailer (Dec 2010 – Dec 2011). The project is unsupervised, aiming at customer segmentation.

Two cleaned tables produced by `data_prep.py` live in `../data/`:

-   `online_retail_clean.csv` – transaction-level (one row per line item)
-   `customer_features.csv` – customer-level RFM + behavioural extras

This notebook does the EDA only; modelling lives in subsequent files.

# 2. Load data

```{r load}
tx <- read_csv("../data/online_retail_clean.csv",
               show_col_types =  FALSE)
customer <- read_csv("../data/customer_features.csv",
                     show_col_types = FALSE)

tx$InvoiceDate <- as_datetime(tx$InvoiceDate)
tx$Date <- as_date(tx$Date)

list(
  transactions = dim(tx),
  customers    = dim(customer),
  date_range   = range(tx$InvoiceDate),
  countries    = n_distinct(tx$Country)
)
```

```{r glimpse-tx}
glimpse(tx)
```

```{r glimpse-customer}
glimpse(customer)
```

# 3. Headline numbers

```{r kpis}
tibble(
  customer = n_distinct(tx$CustomerID),
  invoices = n_distinct(tx$InvoiceNo),
  unique_products = n_distinct(tx$StockCode),
  line_items = nrow(tx),
  total_revenue = sum(tx$Amount),
  avg_basket = mean(tx$Amount)
) |>
  pivot_longer(everything(), names_to = "metric", values_to = "value") |>
  mutate(value = comma(value, accuracy = 0.01))
```

# 4. Distribution of RFM features (raw vs. log)

```{r rfm-raw, fig.height=3.5}
customer |>
  select(Recency, Frequency, Monetary) |>
  pivot_longer(everything(), names_to = "feature", values_to = "value") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 50, fill = "steelblue", colour = "white") +
  facet_wrap(~ feature, scales = "free", nrow = 1) +
  labs(title = "Raw RFM distributions (heavy right tail)",
       x = NULL, y = "Customers")
```

```{r rfm-log, fig.height=3.5}
customer |>
  select(log_Recency, log_Frequency, log_Monetary) |>
  pivot_longer(everything(), names_to = "feature", values_to = "value") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 50, fill = "darkorange", colour = "white") +
  facet_wrap(~ feature, scales = "free", nrow = 1) +
  labs(title = "Log-transformed RFM (much closer to symmetric)",
       x = NULL, y = "Customers")
```

The log scale also makes the wholesale tail in `Monetary` legible -under 
raw scaling those customers are near-invisible, on log scale they form a distinct upper shoulder of the distribution.


# 5. Corraltion structure


```{r corr, fig.height=6}
num_cols <- customer |>
  select(Recency, Frequency, Monetary,
         TotalQuantity, UniqueProducts,
         AvgUnitPrice, AvgBasketValue,
         AvgOrderValue, AvgOrderItems,
         TenureDays, ReturnInvoices, ReturnRate)

corrplot::corrplot(
  cor(num_cols, use = "pairwise.complete.obs"),
  method = "color", type = "upper", order = "hclust",
  addCoef.col = "black", number.cex = 0.7,
  tl.col = "black", tl.cex = 0.8, diag = FALSE
)
```
* `Monetary`, `TotalQuantity`, and `Frequency` are tightly correlated —
  expected, since spending more usually means buying more in more orders.
* `AvgOrderValue` and `AvgBasketValue` carry similar signal; we'll likely
  keep only one in the clustering input.
* `Recency` is largely independent of the spending features — it's its own
  axis of variation. That's why R, F, M is the canonical triple.


# 6. Pairs plot on the modelling set

```{r pairs, fig.height=7, fig.width=8, cache=TRUE}
set.seed(1)
customer |>
  slice_sample(n = 1000) |>
  select(log_Recency, log_Frequency, log_Monetary,
         log_UniqueProducts, log_AvgOrderValue) |>
  GGally::ggpairs(progress = FALSE,
                  lower = list(continuous = wrap("points", alpha = 0.3,
                                                 size = 0.6))) +
  labs(title = "Log-feature pairs plot (1k-customer sample)")
```


# 7. Sales over time

```{r ts-daily}
daily <- tx |>
  group_by(Date) |>
  summarise(revenue = sum(Amount),
            invoices = n_distinct(InvoiceNo), .groups = "drop")

ggplot(daily, aes(Date, revenue)) +
  geom_line(colour = "steelblue") +
  geom_smooth(method = "loess", span = 0.2, se = FALSE,
              colour = "darkorange") +
  scale_y_continuous(labels = label_comma(prefix = "£")) +
  labs(title = "Daily revenue, Dec 2010 – Dec 2011",
       x = NULL, y = "Revenue")
```

The Q4 2011 ramp is the holiday gift season — exactly what we'd expect for
this retailer. Good sanity check that the data makes sense.

```{r ts-dow-hour, fig.height=3.5}
library(gridExtra)

p_dow <- tx |>
  count(DOW = factor(DOW, levels = c("Monday","Tuesday","Wednesday",
                                     "Thursday","Friday","Saturday","Sunday")),
        wt = Amount, name = "revenue") |>
  ggplot(aes(DOW, revenue)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = label_comma(prefix = "£")) +
  labs(title = "Revenue by day of week", x = NULL, y = NULL)

p_hour <- tx |>
  count(Hour, wt = Amount, name = "revenue") |>
  ggplot(aes(Hour, revenue)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = label_comma(prefix = "£")) +
  labs(title = "Revenue by hour of day", x = NULL, y = NULL)

gridExtra::grid.arrange(p_dow, p_hour, nrow = 1)
```

# 8. Geography

```{r country, fig.height=4.5}
tx |>
  count(Country, wt = Amount, name = "revenue", sort = TRUE) |>
  mutate(share = revenue / sum(revenue),
         Country = fct_reorder(Country, revenue)) |>
  slice_max(revenue, n = 10) |>
  ggplot(aes(revenue, Country)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = percent(share, accuracy = 0.1)),
            hjust = -0.1, size = 3) +
  scale_x_continuous(labels = label_comma(prefix = "£"),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Top 10 countries by revenue",
       subtitle = "UK dominance is the headline — Country will be a weak feature",
       x = NULL, y = NULL)
```

# 9. Top products

```{r products}
tx |>
  group_by(StockCode, Description) |>
  summarise(revenue = sum(Amount),
            qty     = sum(Quantity),
            .groups = "drop") |>
  slice_max(revenue, n = 15) |>
  mutate(Description = fct_reorder(Description, revenue)) |>
  ggplot(aes(revenue, Description)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(labels = label_comma(prefix = "£")) +
  labs(title = "Top 15 products by revenue", x = NULL, y = NULL)
```

# 10. Revenue concentration (Pareto check)

How concentrated is revenue across customers? This frames the outlier
question for clustering.

```{r pareto}
pareto <- customer |>
  arrange(desc(Monetary)) |>
  mutate(rank        = row_number(),
         pct_cust    = rank / n(),
         cum_revenue = cumsum(Monetary) / sum(Monetary))

ggplot(pareto, aes(pct_cust, cum_revenue)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_abline(linetype = "dashed", colour = "grey50") +
  scale_x_continuous(labels = percent_format()) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Lorenz curve — customer share vs. revenue share",
       x = "Cumulative share of customers (sorted by spend)",
       y = "Cumulative share of revenue")
```

```{r pareto-numbers}
pareto |>
  filter(pct_cust %in% c(0.01, 0.05, 0.10, 0.20, 0.50)) |>
  transmute(top_pct_customers = percent(pct_cust),
            share_of_revenue  = percent(cum_revenue, accuracy = 0.1))
```

If a tiny share of customers controls a large share of revenue, they'll dominate any unscaled clustering. This confirms the plan to cluster on log-features rather than raw.


# 11. Returns / cancellations

```{r returns, fig.height=3.5}
customer |>
  ggplot(aes(ReturnRate)) +
  geom_histogram(bins = 40, fill = "firebrick", colour = "white") +
  scale_x_continuous(labels = percent_format()) +
  labs(title = "Return rate per customer",
       subtitle = "Most customers never cancel; a small group cancels heavily",
       x = "Cancelled invoices / total invoices", y = "Customers")
```

```{r returns-vs-spend}
ggplot(customer, aes(log_Monetary, ReturnRate)) +
  geom_point(alpha = 0.3, size = 0.8, colour = "steelblue") +
  geom_smooth(method = "loess", se = FALSE, colour = "darkorange") +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Return rate vs. customer spend",
       x = "log(Monetary)", y = "Return rate")
```