---
title: "02_PCA"
author: "Anish"
date: "2026-04-25"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5, dpi = 110              
)

library(tidyverse)
library(scales)

theme_set(theme_minimal(base_size = 12))
```

# 1. Why PCA before clustering

Two distinct jobs PCA does for this project, both straight

1.  **Visulisation** - project the the customers onto PC1-PC2 so we can
    eyeball whether segments separates in the plane of the greatest
    variance, colour by candidate cluster labels later, and produce the
    biplot.
2.  **Preprocessing** - replace the seven correlated input features with
    2-3 orthogonal PC scores and cluster those.

The output of this notebook is stored in PC-scores matrix
('../data/customer_pc_scores.csv') consumed by '03_clustering.Rmd'

# 2. Load the customer-level features

```{r load}
customer <- read_csv("../data/customer_features.csv",
                     show_col_types = FALSE)
glimpse(customer)
```

# 3. Build the modelling matrix

From the EDA (correlation heatmap in `01_EDA.Rmd`) we settled on a
seven-feature input. The redundant ones (`AvgBasketValue` ≈
`AvgOrderValue`, `TotalQuantity` ≈ `Monetary`, `ReturnInvoices` ≈
`Frequency`) were dropped. `Country` was dropped (82.9% UK).
`ReturnRate` is held back as a *profiling* variable — too sparse to
anchor a cluster, but useful for interpretation later.

```{r build-X}
feature_cols <- c("log_Recency", "log_Frequency", "log_Monetary",
                  "log_UniqueProducts", "log_AvgOrderValue",
                  "AvgUnitPrice", "TenureDays")

X <- customer |>
  select(CustomerId, all_of(feature_cols)) |>
  drop_na()

dim(X)
summary(X[, feature_cols])
```

# 4. Fit PCA

`prcomp` with `scale = TRUE` standardises every column to zero mean and
unit variance — required because `AvgUnitPrice` and `TenureDays` are on
totally different scales from the log features.

```{r pca-fit}
pr.out <- prcomp(X[, feature_cols], center = TRUE, scale. = TRUE)
pr.out
```

# 5. Variance explained

```{r pve}
pve <- pr.out$sdev^2 / sum(pr.out$sdev^2)
pve_df <- tibble(
  PC          = factor(paste0("PC", seq_along(pve)),
                       levels = paste0("PC", seq_along(pve))),
  PVE         = pve,
  cumulative  = cumsum(pve)
)
pve_df
```

```{r scree, fig.height=3.5}
p_scree <- ggplot(pve_df, aes(PC, PVE)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = percent(PVE, accuracy = 0.1)),
            vjust = -0.4, size = 3) +
  scale_y_continuous(labels = percent_format(),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Scree plot", x = NULL, y = "Proportion of variance")

p_cum <- ggplot(pve_df, aes(PC, cumulative, group = 1)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 2) +
  geom_hline(yintercept = c(0.7, 0.8, 0.9),
             linetype = "dashed", colour = "grey60") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  labs(title = "Cumulative variance",
       x = NULL, y = "Cumulative PVE")

gridExtra::grid.arrange(p_scree, p_cum, nrow = 1)
```

```{r kaiser}
tibble(
  PC         = paste0("PC", seq_along(pr.out$sdev)),
  eigenvalue = pr.out$sdev^2,
  keep_kaiser = pr.out$sdev^2 > 1
)
```

# 6. Loadings — what each PC actually means

```{r loadings}
loadings <- as_tibble(pr.out$rotation, rownames = "feature")
loadings |> mutate(across(where(is.numeric), \(x) round(x, 2)))
```

```{r loadings-heatmap, fig.height=4}
loadings |>
  pivot_longer(-feature, names_to = "PC", values_to = "loading") |>
  filter(PC %in% paste0("PC", 1:4)) |>
  mutate(PC = factor(PC, levels = paste0("PC", 1:4))) |>
  ggplot(aes(PC, fct_rev(feature), fill = loading)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", loading)), size = 3) +
  scale_fill_gradient2(low = "firebrick", mid = "white",
                       high = "steelblue", midpoint = 0,
                       limits = c(-1, 1)) +
  labs(title = "Loadings (PC1–PC4)", x = NULL, y = NULL)
```

Reading the loadings is what lets us *name* each component. Expected
pattern based on the EDA correlation structure:

-   **PC1** loads positively on `log_Frequency`, `log_Monetary`,
    `log_UniqueProducts`, `log_AvgOrderValue` and negatively on
    `log_Recency` — i.e. an "engagement / customer value" axis.
-   **PC2** is dominated by `TenureDays` and either `AvgUnitPrice` or
    `Recency`'s residual signal — a "tenure / lifecycle" axis orthogonal
    to engagement.
-   **PC3** isolates `AvgUnitPrice` (which the EDA showed was
    near-uncorrelated with everything else) — a "price-point" axis.

# 7. PC scores — the new feature space

```{r scores}
scores <- as_tibble(pr.out$x) |>
  mutate(CustomerId = X$CustomerId, .before = 1) |>
  left_join(customer |> select(CustomerId, Monetary, Frequency,
                               Recency, Country, ReturnRate),
            by = "CustomerId")

head(scores)
```

## 7a. PC1 vs PC2 scatter, coloured by spend

```{r pc12-monetary}
ggplot(scores, aes(PC1, PC2, colour = log10(Monetary + 1))) +
  geom_point(alpha = 0.5, size = 0.8) +
  scale_colour_viridis_c(option = "magma",
                         name = expression(log[10]~Monetary)) +
  labs(title = "Customers in PC1–PC2 space, coloured by spend",
       subtitle = "PC1 should align with engagement/value")
```

## 7b. PC1 vs PC2 scatter, coloured by recency

```{r pc12-recency}
ggplot(scores, aes(PC1, PC2, colour = Recency)) +
  geom_point(alpha = 0.5, size = 0.8) +
  scale_colour_viridis_c(option = "viridis", direction = -1,
                         name = "Recency (days)") +
  labs(title = "PC1–PC2 space, coloured by recency",
       subtitle = "Recent customers should sit on the engagement side of PC1")
```

## 7c. PC2 vs PC3, coloured by AvgUnitPrice

```{r pc23-unit-price}
ggplot(scores, aes(PC2, PC3,
                   colour = pmin(customer$AvgUnitPrice[match(CustomerId,
                                                             customer$CustomerId)],
                                 quantile(customer$AvgUnitPrice, 0.99,
                                          na.rm = TRUE)))) +
  geom_point(alpha = 0.5, size = 0.8) +
  scale_colour_viridis_c(option = "cividis",
                         name = "AvgUnitPrice (capped @ p99)") +
  labs(title = "PC2 vs PC3 — surfacing the price-point axis")
```

# 8. Biplot

That works, but a ggplot version reads more clearly in the report. We
overlay variable arrows (scaled to the score range) on top of the
scatter.

```{r biplot, fig.height=6, fig.width=7}
arrow_scale <- with(pr.out,
                    min(max(abs(x[, 1])) / max(abs(rotation[, 1])),
                        max(abs(x[, 2])) / max(abs(rotation[, 2])))) * 0.7

arrows_df <- as_tibble(pr.out$rotation, rownames = "feature") |>
  mutate(PC1 = PC1 * arrow_scale,
         PC2 = PC2 * arrow_scale)

ggplot(scores, aes(PC1, PC2)) +
  geom_point(alpha = 0.25, size = 0.6, colour = "grey50") +
  geom_segment(data = arrows_df,
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")),
               colour = "firebrick", linewidth = 0.7) +
  geom_text(data = arrows_df,
            aes(x = PC1 * 1.08, y = PC2 * 1.08, label = feature),
            colour = "firebrick", size = 3.5) +
  labs(title = "Biplot — customers + variable loadings on PC1–PC2",
       x = sprintf("PC1 (%s)", percent(pve[1], accuracy = 0.1)),
       y = sprintf("PC2 (%s)", percent(pve[2], accuracy = 0.1)))
```

# 9. Outlier check on PC1

The EDA Pareto curve suggested the top \~1% of customers might pull a
disproportionate amount of variance. With log scaling that should be
much tamer, but worth confirming on PC1.

```{r pc1-outliers}
scores |>
  arrange(desc(PC1)) |>
  slice_head(n = 10) |>
  select(CustomerId, PC1, PC2, Monetary, Frequency, Recency, Country)
```

```{r pc1-hist, fig.height=3.5}
ggplot(scores, aes(PC1)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white") +
  labs(title = "Distribution of PC1 (engagement axis)",
       x = "PC1", y = "Customers")
```

PC1 looks roughly symmetric and unimodal so we're fine to cluster on it
directly.

# 10. Save PC scores for the clustering step

We keep the first **k** PCs where `k` is the smallest number that
crosses 80% cumulative variance — defensible under both the elbow
heuristic and the variance-threshold heuristic. Adjust `k` here if the
scree suggests otherwise.

```{r save-scores}
k <- which(cumsum(pve) >= 0.80)[1]
k

scores_out <- scores |>
  select(CustomerId, all_of(paste0("PC", 1:k)))

write_csv(scores_out, "../data/customer_pc_scores.csv")

list(
  components_kept = k,
  variance_kept   = percent(sum(pve[1:k]), accuracy = 0.1),
  customers       = nrow(scores_out)
)
```
