---
title: "03_clustering"
author: "Anish"
date: "2026-04-26"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5, dpi = 110
)

library(tidyverse)
library(scales)
library(cluster)
library(factoextra)
library(fpc)
library(dendextend)

set.seed(42)
theme_set(theme_minimal(base_size = 12))
```


# 1. Goal of this notebook

# 2. Load PC scores and the original feature table

```{r load}
pc <- read_csv("../data/customer_pc_scores.csv",
               show_col_types = FALSE)
customer <- read_csv("../data/customer_features.csv",
                     show_col_types = FALSE)

dim(pc)
glimpse(pc)
```


```{r matrix}
# Modelling matrix
Xpc <- pc |> select(starts_with("PC")) |> as.matrix()
rownames(Xpc) <- pc$CustomerId
n_pc <- ncol(Xpc)
sprintf("Clustering on %d PCs across %d customers", n_pc, nrow(Xpc))
```


# 3. K-means

## 3a. Choosing K - WCSS / elbow


```{r km-elbow, cache=TRUE}
k_grid <- 2:10
wcss <- map_dbl(k_grid,
                \(k) kmeans(Xpc, centers = k, nstart = 25,
                            iter.max = 50)$tot.withinss)

elbow_df <- tibble(k = k_grid, wcss = wcss)

ggplot(elbow_df, aes(k, wcss)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 2) +
  scale_x_continuous(breaks = k_grid) +
  labs(title = "K-means: within-cluster sum of squares vs k",
       subtitle = "Look for the kink — the elbow",
       x = "k", y = "Total WCSS")
```


## 3b. Silhouette width


```{r km-sil, cache=TRUE}
d <- dist(Xpc)   # used by silhouette and hclust below
sil <- map_dbl(k_grid, \(k) {
  cl <- kmeans(Xpc, centers = k, nstart = 25, iter.max = 50)$cluster
  mean(silhouette(cl, d)[, "sil_width"])
})
sil_df <- tibble(k = k_grid, silhouette = sil)

ggplot(sil_df, aes(k, silhouette)) +
  geom_line(colour = "darkorange", linewidth = 1) +
  geom_point(colour = "darkorange", size = 2) +
  scale_x_continuous(breaks = k_grid) +
  labs(title = "K-means: average silhouette width vs k",
       subtitle = "Higher = more honest separation; pick the peak",
       x = "k", y = "Mean silhouette")
```


## 3c. Gap statistic

The most rigorous of the three (compares within-cluster dispersion to a
uniform-noise reference). Slow — bumping `B` up gives smoother
estimates at the cost of runtime.

```{r km-gap, cache=TRUE}
gap <- clusGap(Xpc, FUN = kmeans, nstart = 25,
               K.max = 10, B = 50)
print(gap, method = "Tibs2001SEmax")

fviz_gap_stat(gap) +
  labs(title = "Gap statistic — Tibshirani 2001 SEmax rule")
```


## 3d. Pick k and fit the final model

The three diagnostics often disagree by ±1. Pick the value that's
defensible under at least two and matches the substantive prior (4–6
clusters from the EDA discussion).

```{r km-final}
k_star <- 3  # CHANGE if your diagnostics point elsewhere

km_fit <- kmeans(Xpc, centers = k_star, nstart = 50, iter.max = 100)
km_fit$size                            # cluster sizes
round(km_fit$centers, 2)               # centroids in PC space

mean(silhouette(km_fit$cluster, d)[, "sil_width"])
```

# 4. Hierarchical clustering

* **Ward (`ward.D2`)** — minimises within-cluster variance, lines up
  philosophically with k-means. Usually the strongest single choice for
  customer segmentation.
* **Complete** — ISLR's default in the lab; tends to give compact,
  similar-diameter clusters.
* **Average** — friendlier to elongated or unequal-density clusters;
  used here as a robustness check.
  

## 4a. Fit and plot dendrograms

```{r hc-fit, cache=TRUE}
hc_ward     <- hclust(d, method = "ward.D2")
hc_complete <- hclust(d, method = "complete")
hc_average  <- hclust(d, method = "average")
```

```{r hc-dendrograms, fig.height=10}
par(mfrow = c(3, 1), mar = c(2, 4, 3, 1))
plot(hc_ward,     labels = FALSE, hang = -1,
     main = "Ward (D2) linkage",     xlab = "", sub = "")
rect.hclust(hc_ward, k = k_star, border = "steelblue")

plot(hc_complete, labels = FALSE, hang = -1,
     main = "Complete linkage",      xlab = "", sub = "")
rect.hclust(hc_complete, k = k_star, border = "darkorange")

plot(hc_average,  labels = FALSE, hang = -1,
     main = "Average linkage",       xlab = "", sub = "")
rect.hclust(hc_average,  k = k_star, border = "firebrick")
par(mfrow = c(1, 1))
```

The dendrogram tells you whether `k_star` is being forced or whether
the tree naturally falls into that many clusters. Ward and complete
should look broadly similar; if average linkage produces one giant
cluster and a handful of singletons, it's flagging a chaining problem
(a real concern with retail data) — that's a finding, not a bug.


## 4b. Cut the tree at k\* and compare to k-means

```{r hc-cut}
hc_cluster <- cutree(hc_ward, k = k_star)

table(kmeans = km_fit$cluster, hclust = hc_cluster)
```

A nearly block-diagonal cross-tab means the two algorithms agree on
who's in which segment (just with different label numbers). Heavy
off-diagonal mass means the algorithms disagree about cluster
boundaries — usually because k-means prefers spherical, equal-size
clusters and hierarchical doesn't.

```{r hc-sil}
mean(silhouette(hc_cluster, d)[, "sil_width"])
```

# 5. Stability — bootstrap Jaccard

This is the validation step that separates "real structure" from
"k-means happened to land somewhere." Resample customers (with
replacement), refit the clusterer, and measure how often the *original*
clusters reappear. Jaccard ≥ 0.75 = "stable cluster", < 0.6 = "probably
noise" (Hennig 2007).

```{r km-stability, cache=TRUE}
boot_km <- clusterboot(
  Xpc,
  B            = 50,
  bootmethod   = "boot",
  clustermethod = kmeansCBI,
  krange       = k_star,
  seed         = 1,
  count        = FALSE
)
round(boot_km$bootmean, 3)            # one Jaccard per cluster
boot_km$bootbrd                        # how often dissolved
```

```{r hc-stability, cache=TRUE}
boot_hc <- clusterboot(
  Xpc,
  B            = 50,
  bootmethod   = "boot",
  clustermethod = hclustCBI,
  method       = "ward.D2",
  k            = k_star,
  seed         = 1,
  count        = FALSE
)
round(boot_hc$bootmean, 3)
boot_hc$bootbrd
```

# 6. Visualise the final clusters in PC1–PC2

```{r km-plot}
plot_df <- pc |>
  mutate(km_cluster = factor(km_fit$cluster),
         hc_cluster = factor(hc_cluster))

ggplot(plot_df, aes(PC1, PC2, colour = km_cluster)) +
  geom_point(alpha = 0.5, size = 0.8) +
  scale_colour_brewer(palette = "Set1", name = "K-means") +
  labs(title = sprintf("K-means clusters (k = %d) in PC1–PC2", k_star))
```

```{r hc-plot}
ggplot(plot_df, aes(PC1, PC2, colour = hc_cluster)) +
  geom_point(alpha = 0.5, size = 0.8) +
  scale_colour_brewer(palette = "Set2", name = "Hierarchical (Ward)") +
  labs(title = sprintf("Hierarchical clusters (Ward, k = %d) in PC1–PC2",
                       k_star))
```


```{r save}
final_label_source <- "kmeans"   # CAN CHANGE to "hclust"

final_cluster <- if (final_label_source == "kmeans")
                   km_fit$cluster else hc_cluster

assignments <- pc |>
  select(CustomerId) |>
  mutate(cluster        = factor(final_cluster),
         cluster_source = final_label_source,
         k              = k_star)

write_csv(assignments, "../data/customer_clusters.csv")

assignments |> count(cluster, name = "customers")
```
